\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-23 paper.tex
More gregor updates

This distinction between graph and local tree arity is mainly
of interest when we consider \emph{unary} nodes, those that have
a single child.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h} are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this region of genome. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{c} has a single child in this
interval and it is unary in this local tree.
% TODO note this is CA but not coalescent, and refer to where we discuss this.
In another example, \noderef{e} is binary in the graph, but is
unary in each of the local trees in which it is present.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
The intervals over which
we know that a node inherits from another, but where there
is no local coalescence, are marked by unary nodes in the
local trees.

Outside of the context of simulation, ARG-unary nodes are
only like to occur in longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-CoV-2; see Discussion).
\emph{Locally} unary nodes, those which have one child
in one or more local trees, are a common and important feature
of many different types of ARG, and a recurring topic
in this article. Without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them.

Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix \ref{sec-cell-lineages-and-args})
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which ARG nodes can be locally unary. Furthermore,
several current ARG inference methods infer locally unary nodes
(see Section~\ref{sec-precision}).

@@

The graph visualisations in Fig.~\ref{fig-simplification} have
some novel features, which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child). Nodes which
are never locally unary therefore have a coalescent span of 100\%,
whereas nodes in which ancestral material never coalesces have a
coalescent span of 0\% (such nodes will be extremely difficult to
identify from the distribution of shared mutations in the samples).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?
% JK: Hmm, not sure we want to jump into inferability here?
All these features will impact what kind of an ARG we can
hope to infer from sampled genomes.

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} genetic sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods based on
(A) parsimony, (B) sampling the SMC$^\prime$, (C) copying from
inferred ancestral haplotypes, and (D) correlated local tree estimation.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
X and Y positions of graph nodes are arbitrary;
line width and node colour are as described in Fig.~\ref{fig-simplification}.
Graphics on the bottom row show the genome positions, relative to the start of the ADH gene,
for each graph edge from the corresponding ARG. Edge intervals are drawn as horizontal lines,
stacked in time order (edges with youngest children at the bottom);
vertical dashed lines indicate breakpoints between local trees.
}
\end{figure}

@@

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args}A,B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples
from the posterior distribution over
ARGs given the data and a fixed model (a discretised version of the SMC$^\prime$ \citep{marjoram2006fast},
an approximation of the full coalescent with recombination).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args}C,D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event is tagged by
a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
occurs at a different location (the ``infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args}C,D have ARG nodes with multiple
parents and multiple children. These indicate deliberate imprecision, where one
or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
Several of these nodes have more than two parents (in red), representing
multiple recombinations inferred to have occurred in an unresolvable order.
Nevertheless, the number of breakpoints in these two ARGs is smaller than the
number of multiple-parent ARG nodes, showing that several multiple-parent nodes
must share breakpoint positions. The sharing of breakpoint locations
between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.
Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
of detail about recombination that lies somewhere between a sequence of unrelated local
trees on one extreme and an ARG with precisely specified recombination events on the other.

@@

% TODO add additional explanation of the bottom row, if enough isn't written below
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
This illustrates that both \tsinfer\ and (to a lesser extent) \relate\ have edges
that span multiple tree boundaries, indicating that they are not inferring
a series of \emph{unrelated} local trees.
%(i.e., with either unlabelled internal
%nodes or differently labelled nodes in each tree).
For example, both infer that \noderef{Af-f}, \noderef{Fr-f} and \noderef{Wa-f}
share a common ancestor along
the entire sequence. However, in comparison to \kwarg\ and \argweaver\ neither
method results in extensive node sharing in the
oldest time periods. This primarily reflects the fact that they do not encode topology
above the local MRCAs, also indicated by multiple roots in the ARG
(nodes with zero parents, in blue), compared to the
single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
\tsinfer\ and \relate\ show a more pragmatic representation: we do not expect
genetic variation to be very informative about topological relationships
between nodes that that are older than the local MRCAs.

@@

\section{Discussion}
\label{sec-discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in development of ARG-based inference methods, their evaluation
and application.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix~\ref{sec-arg-history}, the term
``ancestral recombination graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definition
within the context of stochastic
processes and instead use the term as a more general representation of any
recombinant genetic ancestry seems useful, and we have
tried to clarify and systematise it here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG (gARG) encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which is more limited in what can be described, because of its origins in an idealised mathematical model.

@@

The timing, positions, and even the number of recombination events is generally
not possible to infer precisely from genome sequencing data. For ARGs under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region. For general
ARGs, for instance those representing pedigrees, or genealogies of organisms with
low mutation rates and complex patterns of recombination, such inference can be
even more challenging. The fact that the eARG encoding \emph{requires}
precise estimates about recombination is therefore a fundamental limitation, which
is a central message of this paper.
% This might be where we mention Fisher's junctions and IBD?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121

>>>>

This distinction between graph and local tree arity is mainly
of interest when we consider \emph{unary} nodes, those that have
a single child.
Returning to the example in Fig~\ref{fig-arg-in-pedigree}, nodes
\noderef{g} and \noderef{h} are ARG-unary (Fig~\ref{fig-arg-in-pedigree}B), and are consequently
also unary in the local trees (Fig~\ref{fig-arg-in-pedigree}C).
On the other hand, node \noderef{f} has two children
in the graph, but is binary only
in the local tree covering the interval $[2, 7)$,
representing the coalescence of samples \noderef{a} and \noderef{c}
in this genome region. Over the interval $[0, 2)$ no coalescence occurs,
but we still record the fact that genome \noderef{c} inherits from \noderef{f}
in the local tree. Thus, \noderef{c} has a single child in this
interval and it is unary in this local tree.
% TODO note this is CA but not coalescent, and refer to where we discuss this.
In another example, \noderef{e} is binary in the graph, but is
unary in each of the local trees in which it is present.
This is because \noderef{a} and \noderef{c} are both recombinant
offspring of \noderef{e} and \noderef{f}, and,
in this example,
\noderef{a} inherits from
\noderef{e} over the interval $[0, 2)$,
and \noderef{c} inherits from \noderef{e} over $[7, 10)$.
The intervals over which
we know that a node inherits from another, but where there
is no local coalescence, are marked by unary nodes in the
local trees.

Outside of the context of simulation, ARG-unary nodes are
only likely to occur in longitudinal datasets where genetic
data is sampled at many timepoints and recombination
is rare (e.g.~SARS-CoV-2; see Discussion).
\emph{Locally} unary nodes, those which have one child
in one or more local trees, are a common and important feature
of many different types of ARG, and a recurring topic
in this article. Without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them.

% FIXME: move to top of section
Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix \ref{sec-cell-lineages-and-args})
and there is usually
little information about these intermediates, and hence limited utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which ARG nodes can be locally unary. Furthermore,
several current ARG inference methods infer locally unary nodes
(see Section~\ref{sec-precision}).

@@

The graph visualisations in Fig.~\ref{fig-simplification} have
three novel features.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the ``coalescent span'', the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child). Nodes which
are never locally unary therefore have a coalescent span of 100\%,
whereas nodes in which ancestral material never coalesces have a
coalescent span of 0\% (such nodes will be extremely difficult to
identify from the distribution of shared mutations in the samples).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?
% JK: Hmm, not sure we want to jump into inferability here?
All these features will impact what kind of an ARG we can
hope to infer from sampled genomes.

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods based on
% FIXME: this is the wrong place to be summarising. Name the methods
% here and summarise in the text
(A) parsimony, (B) sampling the SMC$^\prime$, (C) copying from
inferred ancestral haplotypes, and (D) correlated local tree estimation.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
X and Y positions of graph nodes are arbitrary;
line width and node colour are as described in Fig.~\ref{fig-simplification}.
Graphics on the bottom row show the genome positions, relative to the start of the ADH gene,
for each graph edge from the corresponding ARG. Edge intervals are drawn as horizontal lines,
stacked in time order (edges with youngest children at the bottom);
vertical dashed lines indicate breakpoints between local trees.
}
\end{figure}

@@

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args}A,B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples
from the posterior distribution over
ARGs given the data and a fixed model (a discretised version of the SMC$^\prime$ \citep{marjoram2006fast},
an approximation of the full coalescent with recombination).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args}C,D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is an important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

% Here we describe recombination features
A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event is tagged by
a node with exactly 2 parents and one child. Assuming that the breakpoint for each event
% TODO "infinite breakpoints" is goign to confuse people. What are we
% clarifying here? Citation necessary if we keep.
occurs at a different location (the ``infinite breakpoints'' model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args}C,D have ARG nodes with multiple
parents and multiple children. These indicate deliberate imprecision, where one
or more recombinations have been ``pushed down'' onto a more recent, identifiable node.
Several of these nodes have more than two parents (in red), representing
multiple recombinations inferred to have occurred in an unresolvable order.
Nevertheless, the number of breakpoints in these two ARGs is smaller than the
number of multiple-parent ARG nodes, showing that several multiple-parent nodes
must share breakpoint positions. The sharing of breakpoint locations
between these nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.
Overall, Fig.~\ref{fig-inferred-args} shows that \tsinfer\ and \relate\ ARGs contain a level
of detail about recombination that lies somewhere between a sequence of unrelated local
trees on one extreme and an ARG with precisely specified recombination events on the other.

@@

% TODO add additional explanation of the bottom row, if enough isn't written below
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
This illustrates that both \tsinfer\ and (to a lesser extent) \relate\ have edges
that span multiple tree boundaries, indicating that they are not inferring
a series of \emph{unrelated} local trees.
%(i.e., with either unlabelled internal
%nodes or differently labelled nodes in each tree).
For example, both infer that \noderef{Af-f}, \noderef{Fr-f}, and \noderef{Wa-f}
share a common ancestor along
the entire sequence. However, in comparison to \kwarg\ and \argweaver\ neither
method results in extensive node sharing in the
oldest time periods. This primarily reflects the fact that they do not encode topology
above the local MRCAs, also indicated by multiple roots in the ARG
(nodes with zero parents, in blue), compared to the
single ARG root in the \kwarg\ and \argweaver\ genealogies. In terms of inference,
\tsinfer\ and \relate\ show a pragmatic representation: we do not expect
genetic variation to be very informative about topological relationships
between nodes that that are older than the local MRCAs.

@@

\section{Discussion}
\label{sec-discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in development of ARG-based inference methods, their evaluation
and application.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix~\ref{sec-arg-history}, the term
``ancestral recombination graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definition
within the context of stochastic
processes and instead use the term as a more general representation of any
recombinant genetic ancestry seems useful, and we have
tried to clarify and systematise it here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG (gARG) encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which is more limited in what can be described,
because of its origins in an idealised mathematical model,
and the assumption of having complete information about
recombination events.

@@

The timing, positions, and even the number of recombination events is generally
not possible to infer precisely from genome sequencing data. For ARGs under
coalescent-based models, the proportion of recombination events that change the
ARG topology grows very slowly with sample size \citep{hein2004gene}, and of those
events only a small proportion are actually detectable from the data, assuming
human-like mutation rates \citep{myers2002detection,hayman2023recoverability}.
Even when a recombination event \emph{is} detectable, its timing and breakpoint
position can only be inferred approximately, depending on how much information
can be elucidated from mutations in the surrounding genomic region. For general
ARGs, for instance those representing pedigrees, or genealogies of organisms with
low mutation rates and complex patterns of recombination, such inference can be
even more challenging. The fact that the eARG encoding \emph{requires}
precise information about recombination is therefore a fundamental limitation, which
is a central message of this paper.
% This might be where we mention Fisher's junctions and IBD?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121
